\Large \bf \Toolname: Cache Line Contention Detection for COTS Binaries

### 2014-12-18 HotOS'15/paper.tex
I think this is good enough for now.

% For leaving some comments in the draft.
\newcommand{\comment}[1]{}
\newcommand{\TextToolname}{Malcontent}
\newcommand{\Toolname}{\textsc{\TextToolname{}}}
\newcommand{\BoldToolname}{\textsc{\bf\TextToolname{}}}

@@

% --------------------------------------------------------------------------------------
\section{Methodology}\label{sec:methodology}

@@

% --------------------------------------
\subsection{Training}

% --------------------------------------
\subsection{Shadow Memory}
\begin{figure*}
\includegraphics[width=\textwidth]{shadow.pdf}
\caption{\label{fig:layout_of_shadow}Layout of shadow memory. Native cache lines map to individual shadow state cells. %
 In the above %
diagram, the cache line associated with bytes [\texttt{0xF00},\texttt{0xF40}) are being sampled using $SampleId=11$. When %
we look at entry $11$ in the sample table, we see that the monitor thread chose this sample using a recent allocation of an %
object with $TypeId=4$.}
\end{figure*}

@@

% --------------------------------------
\subsection{Instrumentation}
If a sampled memory location is accessed, then the instrumentation instructions of \Figref{inline_assembly} will detect the non-zero
$SampleId$, and invoke the $CheckContention$ function if the current thread does not already appear to own the sampled cache line.
The $CheckContention$ algorithm (\Figref{algorithm}) is responsible for acquiring ownership of a cache line, as well as detecting
contention. At a high level, the algorithm detects contention by failing to acquire ownership. The two key differences between the
$CheckContention$ algorithm and prior work are: \begin{enumerate}
	\item The owner thread can be \emph{optionally} paused. Pausing the owner thread emulates the race-detection capability of
DataCollider. Using this approach, we can pinpoint two locations where contention (and potentially races) can
absolutely occur.

Not pausing the owner thread means that \Toolname{} will detect contention as an ownership transfer. The benefit of
this approach is that it can find contention even within critical sections. The drawback is that accuracy will not be perfect: the last access of
the sampled cache line by the owner thread might not be the instruction at which it obtained ownership of the cache line.

@@

% TODO: Measure these improvements!!
	\item Unlike in Dr. Memory, sampling (and therefore detection) is immediately disabled once contention is detected. This improves
performance in the contended case because it reduces contention on the shadow memory itself.
\end{enumerate}

% --------------------------------------
\subsection{Reporting}

@@

\begin{figure}[t]
\lstset{language=[x64]Assembler}
\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
;; Racy watchpoint check.
CMP [Shadow(Addr) + Offs(SampleId)], 0
JZ ignore
;; Racy ownership check.
CMP [Shadow(Addr) + Offs(ThreadId)], MyThreadId
JZ ignore
;; Take ownership, or transfer ownership
CALL CheckContention(Addr,PC)
ignore:
\end{lstlisting}
\caption{\label{fig:inline_assembly}Instrumentation instructions injected for each shared memory access. The algorithm %
optimizes on instruction footprint by performing lightweight, racy checks up front as a way of avoiding more heavyweight %
contention detection. If the cache line appears watched, and if the current thread doesn't own the line, then the %
$CheckContention$ algorithm is invoked. This algorithm either acquires ownership for the current thread, or detects and %
reports contention.}
\end{figure}

@@

%\item We iteratively refine the granularity of proxy memory, starting from the page granularity and going down to the cache

@@

% --------------------------------------------------------------------------------------
\section{Evaluation}\label{sec:evaluation}
Our evaluation aims to answer the following questions:
\begin{itemize}
\item Does the training phase significant reduce the performance overheads of the detection phase?

\end{itemize}

>>>>

% Name of the tool.
\newcommand{\TextToolname}{Malcontent}
\newcommand{\Toolname}{\textsc{\TextToolname{}}}
\newcommand{\BoldToolname}{\textsc{\bf\TextToolname{}}}

\renewcommand{\Secref}[1]{Section~\ref{sec:#1}}

@@

% --------------------------------------------------------------------------------------
\section{Implementation}\label{sec:implementation}

@@

% --------------------------------------
\subsection{Interposition}

\begin{figure}[t]
\begin{lstlisting}[language=C,basicstyle=\footnotesize\ttfamily]
WRAP_FUNCTION(libc, malloc, (void *),
                            (size_t size)) {
  auto malloc = WRAPPED_FUNCTION;
  auto context = RETURN_ADDRESS;
  auto addr = malloc(size);
  if (addr) {
    auto type_id = TypeIdFor(context, size);
    AddSamplePoint(type_id, addr);
  }
  return addr;
}
\end{lstlisting}
\caption{\label{fig:wrapper}Wrapper for the \texttt{malloc} function in \texttt{libc}. The wrapper invokes the native malloc %
function, thus preserving native memory layouts, and determines the $TypeId$ for the allocated address. Finally, the }
\end{figure}

The interposition layer is responsible for assigning $TypeId$s to different types of allocated objects, and maintaining lists of
the most recently allocated object addresses on a per-type basis. These lists are used by the monitor to comprehensively
sample the heap.

\Toolname{} uses Granary's wrapper mechanism \cite{Granary} to interpose on the dynamic memory allocation routines
used by the application. By default, \Toolname{} interposes on \texttt{malloc} and other POSIX memory management
routines  available from \texttt{libc}, as well as \texttt{new} from \texttt{libc++}. \Toolname{} can also be configured to
interpose on non-standard allocators, either by using symbol information, or the addresses of the allocators within the
binaries to be instrumented.

\Figref{wrapper} shows an example wrapper for the \texttt{malloc} function. The wrapper determines the $TypeId$ for the
allocated object given the size of the object and the allocation context (heuristically estimated using the return address into
the caller of the \texttt{malloc} function). For finer-grained $TypeId$ associations, or more detailed reporting, calls stacks
can be provided as the allocation context.

{\bf Future work:} Like SHERIFF \cite{SHERIFF}, try to fix some cache contention issues by padding allocated objects.

% --------------------------------------
\subsection{Monitor}

The monitor thread is spawned at program startup and is responsible for periodically choosing memory locations to sample,
then enabling their watchpoints. When a sampling period is over, the enabled watchpoints are disabled, and any detected
contention is reported.

The monitor employs a round-robin sampling algorithm. For each $TypeId$, it chooses the most recently allocated object, as
recorded by the interposition layer. Then, it enables \emph{all} watchpoints associated with the cache lines spanned by
the allocated object. When the maximum configured number of samples have been placed, it goes to sleep, so as to give the
placed watchpoints time to be hit and detect contention. Eventually, it wakes up and reported any instances of detected
contention.

{\bf Future work: Biased sampling.} Bias sampling toward objects that are more typically accessed in \emph{cold} code.
This would agree with the ``cold code hypothesis'' \cite{LiteRace}, wherein important sources of cache contention (in
the case of \cite{LiteRace}, the focus was data races) are likely to be in colder code because otherwise those issues would
have already been discovered, or are benign.

{\bf Future work: Threshold sampling.} Disable sampling of some types when the number of detected issues exceeds some
bound.

% --------------------------------------
\subsection{Instrumentation}
\paragraph{Shadow Memory}
\begin{figure*}
\includegraphics[width=\textwidth]{shadow.pdf}
\caption{\label{fig:layout_of_shadow}Layout of shadow memory. Native cache lines map to individual shadow state cells. %
 In the above %
diagram, the cache line associated with bytes [\texttt{0xF00},\texttt{0xF40}) are being sampled using $SampleId=11$. When %
we look at entry $11$ in the sample table, we see that the monitor thread chose this sample using a recent allocation of an %
object with $TypeId=4$.}
\end{figure*}

@@

\paragraph{Shadow Instrumentation}

Shadow memory instrumentation is applied only to those memory-accessing instructions identified during the training phase. For all
other instructions, Granary's ``null" instrumentation is applied. In some cases, \Toolname{} can even instruct Granary to execute code
natively. The rest of this section focuses on shadow instrumentation.

If a sampled memory location is accessed, then the instrumentation instructions of \Figref{inline_assembly} will detect the non-zero
$SampleId$, and invoke the $CheckContention$ function if the current thread does not already appear to own the sampled cache line.
The $CheckContention$ algorithm (\Figref{algorithm}) is responsible for acquiring ownership of a cache line, as well as detecting
contention. At a high level, the algorithm detects contention by failing to acquire ownership. The two key differences between the
$CheckContention$ algorithm and prior work are: \begin{enumerate}
	\item The owner thread can be \emph{optionally} paused. Pausing the owner thread emulates the race-detection capability of
DataCollider. Using this approach, we can pinpoint two locations where contention (and potentially races) can
absolutely occur.

Not pausing the owner thread means that \Toolname{} can detect contention even within critical sections. The drawback
is that accuracy will not be perfect: the last access of the sampled cache line by the owner thread might not be the
instruction at which it obtained ownership of the cache line.

@@

% TODO: Measure these improvements!!
	\item Unlike in Dr. Memory, sampling (and therefore detection) is immediately disabled once contention is detected. This improves
performance in the contended case because it reduces contention on the cache lines associated with shadow memory, as well as reduces
CPU overheads due to executing fewer inject instrumentation instructions.
\end{enumerate}

@@

\begin{figure}
\lstset{language=[x64]Assembler}
\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
;; Racy watchpoint check.
CMP [Shadow(Addr) + Offs(SampleId)], 0
JZ ignore
;; Racy ownership check.
CMP [Shadow(Addr) + Offs(ThreadId)], MyThreadId
JZ ignore
;; Take ownership, or transfer ownership
CALL CheckContention(Addr,PC)
ignore:
\end{lstlisting}
\caption{\label{fig:inline_assembly}Instrumentation instructions injected for each shared memory access. The algorithm %
optimizes on instruction footprint by performing lightweight, racy checks up front as a way of avoiding more heavyweight %
contention detection. If the cache line appears watched, and if the current thread doesn't own the line, then the %
$CheckContention$ algorithm is invoked. This algorithm either acquires ownership for the current thread, or detects and %
reports contention.}
\end{figure}

@@

%\item We iteratively refine the granularity of proxy memory, starting from the page granularity and goig down to the cache

@@

% --------------------------------------------------------------------------------------
\section{Evaluation}\label{sec:evaluation}
I want to evaluate the following things:

\paragraph{Training} Here, I want to focus on measuring the reduction in overheads provided by the training phase. There will
be two ways of showing this. The first is tabular, and will compare dynamic instruction counts. Specifically, it will measure how
many instructions would execute with heavy instrumentation with training, how many without, and how many execute natively.

Another thing that might be worth showing is that for some levels of sampling (e.g. 16 samples, 64 samples, 356 samples, etc.) how
many of the heavily instrumented instructions go through a "heavyweight" instrumentation path. This could give insights on
the actual coverage of sampling, and might inform us on whether or not we need to change our sampling strategies. Hopefully
there will be some insights there.

The next is way of showing the benefits are in terms of overhead. Here, we could either focus on a particular benchmark (e.g.
Phoenix, PARSEC), or on a few applications (MySQL, Memcached, etc.). This would be a standard grouped bar chart, with three
bars per application. Each bar is overhead relative to native. The first bar for each application shows baseline Granary overheads.
The second bar shows overheads without training. The third bar shows overheads with training.

\paragraph{Sampling}
Here the purpose is to compare the performance benefits of ``comprehensive sampling" vs. ``comprehensive tracking". This serves
as a direct comparison to Dr. Memory. Again, we want a grouped bar chart. The first bar for each application is a Granary version
of Dr. Memory, which is a basically our tool without training without sampling. The second group of bars is for untrained \Toolname{}
but using sampling, with increasing number of watchpoints. The third group of bars is for trained, with the same numbers of watchpoints.

It would also be nice if we could have a small table that chooses a select few of the results from the graph, and shows data cache
hit/miss ratios, as well as icache hit/miss ratios.

\paragraph{Memory Overheads}
This needs to show the memory overheads of \Toolname{}, both in terms of shadow and in terms of code cache. I've described that
the shadow has an upper-bound on its memory overheads; however, I haven't actually implemented it as such. In practice though,
that might not matter if none of our tests reach the theoretical upper bound.

\paragraph{Effectiveness}
For this to be useful, we need to find real bugs, and showing an example of a report for a real bug will be key I think. In the short
time before HotOS, I don't expect to find much, as getting everything working will be its own challenge. This isn't discouraging me
though as I feel that this is the first ``real" tool that Granary has, and for that reason, it's something that I'm very pleased with.

\section{Conclusion}
This paper presented \Toolname{}, a new cache contention detection framework that combines the strengths of DataCollider and
Dr. Memory into a unified framework. The key novelties of \Toolname{} are the separation of the problem into two phases. The
first phase figures out what code can contain bugs. The second phase detects bugs in that code. The tool, but more specifically
Granary itself, is still under heavy development, and so performance and effectiveness evaluations could not be performed. It is
hoped that Granary's stability will improve enough in time to get some kind of numbers.
